Querying Integrated Scientific Observation and Measurement Data

### 2010-05-10 docs/algdraft/DataIntegration/DataIntegration.tex
(1) Refine description of query rewriting (Sec IV.A) (except the examples) (2) Change some inconsistent symboles in Section III (model) and Sec IV.B (materialization) (3) Refine the first 4 paragraphs of Section IV.B.

For data table $d$ with annotation $A$, the system keeps the
information $AD$ = $\{(d_{id}, A_{id}, d_{meta})\}$.
As we described before, we have four main objects to describe: 
observation type (OT), measurement type (MT), and
context type (CT), and the mapping (Map) from the measurement type to
resource attributes.  
So, the annotation contains the following information.

\begin{itemize}
\item $OT$ = $\{(\underline{A_{id},ot_{id}}, et, isDistinct)\}$ to
  describe an observation type, which denotes on which
  entity type (object in the real world) the observation is made. 
Very often, more than one observation can be made on one entity. Thus,{\em
  isDistinct} is used to denote whether the same value of key
meaasurements of an obbservation types can uniquely identify one
observation or not. 
\item $MT$ = $\{(\underline{A_{id}, mt_{id}}, ot_{id},isKey,
  Cha, \cdots)\}$. Generally, $MT$ contains information about
  characteristic ($Cha$), Standard, Protocol and Precision. We do not include them
  here to make the description clearer. Here, {\em isKey} is used to denote
  whether one measurement type is the key measurement for the
  observation type $OT_{id}$ or not. 
\item $CT$ = $\{(\underline{A_{id}, ct_{id}}, ot_{id}, cot_{id}, Rel,
  isIdentify)\}$ where $cot$ represents the context observation
  type. 
\item $Map$ = $\{(\underline{A_{id},mt_{id}, attr_i, MapCond},
  Val\}$
\end{itemize}

@@

%\begin{example}
\begin{itemize}
\item $Observation: \langle Species='Picea~rubens'\rangle$
\item $Observation_1: \langle Species='Picea~rubens'\rangle$\\
  $\wedge IN(Observation_1, Observation_2$) \\
$\wedge Observation_2: \langle Meas_{unknown}=`California'\rangle$
\item $Observation: \langle Species='Picea~rubens'$\\
$\wedge count(distinct~Species)\geq 5\rangle$
\item $Observation_1: \langle avg(distinct~tree.height)\geq 20.0 \rangle$ 
$\wedge IN(Observation_1, Observation_2)$\\
$\wedge Observation_2: \langle State='California'\rangle$\\
%Give me the data sets that have trees with average ``height''
%  than 20.0 in ``California''
\end{itemize}

@@

As described in Section \ref{sec:dataquery}, 
an OM query is represented using 
OM concept terminologies such as observation
measurement, characteristic, etc. 
To answer such query from the original data model (i.e., dataset
metadata and data tables), we need to {\em rewrite} the quey over the
real data table. 
In what follows, we first sketch the re-writing process. Then, we
detailed the procedure in using the data model and
structures. Finally, we include the process for the more complicated
cases (e.g., with distinct, aggregate).

Roughly speaking, the query rewriting consists of two steps:
\begin{itemize}
\item From the given query, find the relevant data tables and
  attributes that need to be used answer the given query. 
\item Translate the given query to queries over the relevant data
  tables. 
\end{itemize}

@@

The first step is to map the OM query to the real data structure. 
We can utilize the annotation structure $A.Map$ which keeps the correspondences between
measurement type and table attributes. 
So, when the given concept is {\em measurement}, we can directly get
its related attribute names and annotation id $A_{id}$. 
In case that the given OM concept is a non-measurement concept, we can use
other Annotation structures $A.*$ to figure out the measurementt types. 
E.g., if the OM concept is Characteristic ``Species'', we can use
$A.MeasType$ to get the measurement type id.

As mentioned in Section \ref{sec:annotation}, when some annotations are done over data tables, the system keeps the
information $\langle A_{id}, DTableId\rangle$. 
With the $A_{id}$, we can ge the data table information from the
metadata  $\langle A_{id}, DTableId\rangle$. 
Then, we get the relevant data table information and the needed attribute
information.

Let $RD$ be the set of relevant data tables and $d_{a_1} \cdots d_{a_m}$
are the related attributes for table $d\in RD$. We can translate the original OM query to SQL queries
over the data tables.

@@

The above describes the process for the most basic OM query. 
However, in the real application, people ask queries with distinct
observation or entity constraints. Also, people ask queries about
different observatitons using context.

%Discuss how to deal with the distinct constraints.
Many times, a user may want to find {\em distinct} observations or
entities. In this case, the above simple solution cannot work
correctly because it treats observations in different rows as the same
observation. So, we need to use the annotation to figure out what
distinguishes one observation from another. 
When an observation type is denoted with {\em distinct yes} and some
of its measurement types are denoted with {\em key yes}, 
the values of the measurements on the key measurement types can
uniquely distinguish one observation from the other. 
To implement this, in SQL, we can perform ``group by'' operations on
the key measurement types of the observations.

%Discuss how to deal with context
The above description discusses cases without context. 
However, in real application, the observation has context. 
When the context is not {\em identifying}, i.e., the context value
does not affect the uniqueness of one observation, this can be
processed as the basic way. 
When the context is denoted with {\em identifying yes}, it means the
identity of one observation also depend on the contextual
observation. 
In this case, we need to figure out the all the key measurement types
of one observation, then the ``GROUP BY'' operation should be on all
these key measurement types.

{\bf TODO: put an example here.}

In this case, there is no need to change the SQL scripts.
the only change are the relevant attribute set $d_{a_1} \cdots
d_{a_m}$, which corresponds to not only the measurement types that
directly related to the observation, but also the measurement types
that are in the identifying context chain of the observation.

\begin{example}\label{eg:query_reqerite_q1}
{\bf TODO: refine this example} 
Take $Q_1$ as example. The {\tt concept} is ``observation'' and {\tt cond} is {\em species = ``Picea rubens''};

From the annotation $A.MeasType$, we can find the measurement type
$m_3$ with the characteristic {\em species}. 
From the mapping $A.Map$, we further find that the attribute {\em spp} that the
measurement type $m_3$ is mapped to.

The second step is to find the data file which really have this value. 
For this one, we find the data table, and do a selection on the table
content.

{\bf TODO: SQL here.} 
\end{example}

\from{HP}{Refine the process of $Q_2$.\\
For the first step, we find the data tables that contain the needed concepts. 
In each table, we need to figure out the distinct observations. 
From $m_i$ $obstype_i$, all the other measurements that are of
$obstype_i$. 
Group by the key measurements of $obstype_i$ if distinct yes.
If this observation type does not have any context, then the key
measurements are all the measurements directly defined under this
observatioin type. Otherwise, i.e., this observation type has context,
then the key measurements are all the measurements in the context
chain. }

%Analysis of this method.

{\bf Cost analysis:} 
From the above description, we can tell that the computation cost is to search the different data
tables. This way, for each candidate data table, we need to send an
SQL query to the server for evaluation purpose.
If the number of candidate data tables are small. This method should
work well. However, when we have a lot of candidate data tables, the
efficiency may be affected. 
In what follows, we introduce another strategy, which logically merge
the data content in different tables and perform queries ober the
materialized database.

@@

The previous section shows the method to rewrite a given query. 
However, as we analyzed, the computation cost may be higher when the
candidate data tables are many. In this section, we propose another
strategy to make use of the existing optimization strategies of
current DBMS.

The basic idea is to materialize the data into some centralized
concept tables with instances of entities, observations, measurements
and the contexts between observations. 
We call such concept tables core OBOE tables, and denote them as
$OBOE.*$. 
In particular, we have the following concept instance tables. 
\begin{itemize}
%\item $EI$ = $\{(\underline{ei_{id}}, et_{id})\}$ for all the entity instances;
\item $OI$ = $\{(\underline{oi_{id}}, ot_{id}, ei_{id})\}$ keeps
  all the observation instances where $ot_{id}$ is used to link to the
  annotation information in $OT$; 
\item $MI$ = $\{(\underline{mi_{id}}, oi_{id}, mt_{id},
  mVal)\}$ for all the measurement instances where $mt_{id}$ is used to link to the
  annotation information in $MT$; 
\item $CI =\{(\underline{oi_{id},coi_{id},ct_{id}})\}$ for
  all the context instances where $ct_{id}$ is used to link to the
  annotation information in $CT$. 
\end{itemize}

In what follows, we describe how to materialize the data into such
concept instance tables in Section \ref{sec:materialilzedb} when
considering all the distinct and context constratins. 
Then, in Section \ref{sec:query_materialilzedb}, we describe our strategy to
perform queries over materialized database.

\subsubsection{Materialize data}\label{sec:materialilzedb}
\from{HP}{TODO: refine the writing in MaterializeDB and move them here.}

@@

Once the data are materialized using the annotation constraints, we
can perform the query over them. In what follows, we describe how to
use the materialized data to answer the query.

@@

\vspace{0.1in}
{\tt 
\begin{tabular}{l}
SELECT distinct $AD.d_{id}$\\
FROM $AD, MI~as~mi, MT~as~mt$\\
WHERE $mt.Cha='Species'$ \\
$\qquad\quad$ AND $mi.mVal = `Picea rubens'$\\
$\qquad\quad$ AND $mi.mt_{id}=mt.mt_{id}$\\
$\qquad\quad$ AND $mt.A_{id}=AD.A_{id}$\\
GROUP BY $d_{id},oi_{id}$\\
HAVING COUNT(*)$>$5; 
}
\end{tabular}
}
\vspace{0.1in}
\end{example}

>>>>

For a data table $d$ with annotation $A$, the system keeps the
corresponding information between them in $AD$ tables. 
Besides it, we also have four main concepts to describe: 
observation type (OT), measurement type (MT), and
context type (CT), and the mapping (Map) from the measurement type to
resource attributes.  
So, the annotation contains the following information.

\begin{itemize}
\item $AD$ = $\{(d_{id}, A_{id}, d_{meta})\}$ to keep the information
  of data sets $d_{id}$-s and their related annotations
  $A_{id}$. Here, $d_{meta}$ is an abstract attribute for all the
  other metadata of $d_{id}$. 
\item $OT$ = $\{(\underline{A_{id},ot_{id}}, et, isDistinct)\}$ to
  describe an observation type. It denotes on which
  entity type (object in the real world) the observation is made. 
Very often, more than one observation can be made on one entity. Thus,{\em
  isDistinct} is used to denote whether the same value of key
measurements of an obbservation types can uniquely identify one
observation or not. 
\item $MT$ = $\{(\underline{A_{id}, mt_{id}}, ot_{id},isKey,
  Cha, \cdots)\}$. Generally, $MT$ contains information about
  characteristic ($Cha$), Standard, Protocol and Precision. We do not include them
  here to make the description clearer. Here, {\em isKey} is used to denote
  whether one measurement type is the key measurement for the
  observation type $OT_{id}$ or not. 
\item $CT$ = $\{(\underline{A_{id}, ct_{id}}, ot_{id}, cot_{id}, Rel,
  isIdentify)\}$ where $cot$ represents the context observation
  type. 
\item $Map$ = $\{(\underline{A_{id},mt_{id}, attr_i, mapCond, mapVal},
  Val\}$. This structure denotes the correspondences between an
  measurement type ($mt_{id}$ and an attribute $attr_i$ in an dataset
  $d_{id}$ for a given condition $mapCond$. 
\end{itemize}

@@

%\begin{example}
\begin{itemize}
\item $Q_1$ = $Observation: \langle Species='Picea~rubens'\rangle$
\item $Q_2$ = $Observation_1: \langle Species='Picea~rubens'\rangle$\\
  $\wedge IN(Observation_1, Observation_2$) \\
$\wedge Observation_2: \langle Meas_{unknown}=`California'\rangle$
\item $Q_3$ = $Observation: \langle Species='Picea~rubens'$\\
$\wedge count(distinct~Species)\geq 5\rangle$
\item $Q_4$ = $Observation_1: \langle avg(distinct~tree.height)\geq 20.0 \rangle$ 
$\wedge IN(Observation_1, Observation_2)$\\
$\wedge Observation_2: \langle State='California'\rangle$\\
%Give me the data sets that have trees with average ``height''
%  than 20.0 in ``California''
\end{itemize}

@@

As described in Section \ref{sec:dataquery}, 
an OM query is represented using 
OM concept terminologies such as observation, 
measurement, characteristic, etc. 
To answer such queries from the original data model (i.e., dataset 
metadata and translated data tables), 
we need to {\em rewrite} the OM query over the real data tables. 
In what follows, we first sketch the re-writing process. 
Then, we detail the procedure in using the data model and structures. 
Finally, we include the process for the more complicated
cases (e.g., with distinct, aggregate).

%Framework of query rewriting
Roughly speaking, the query rewriting consists of two steps:
\begin{itemize}
\item From the given query, search the database metadata and
  annotation information to find the relevant data tables and
  attributes that need to be used to answer the given query. 
\item Translate the given OM query to queries over the relevant data
  tables. 
\end{itemize}

@@

%Elaborate the first step
The first step is to map the OM concepts in the query to the real data
structure. 
With the annotation structures $OT$, $CT$, and $MT$, any given OM
concepts can be linked to a measurement type $mt_{id}$. 
Then, the annotation structure $Map$ can be used to bridge OM concepts and data sets. 
Utilizing $Map$, we can find the attributes $attr_i$ and the
annotation $A_{id}$).
E.g., when the given concept is {\em measurement}, we can directly get
its related attribute names and annotation id $A_{id}$. 
In case that the given OM concept is a non-measurement concept, we can use
other annotation structures to figure out the measurementt types. 
E.g., if the OM concept is {\em characteristic} ``Species'', we can use
$MT$ to get the measurement type id $mt_{id}$. 
After we get $A_{id}$, we can quickly find the dataset id $d_{id}$
using $AD$ metadata table (Section \ref{sec:annotation}). 
Then, we get the relevant data tables and the needed attributes.

%Elaborate the second step
Let $D_{cand}$ be the set of relevant (or candidate) data tables and $d_{a_1} \cdots d_{a_m}$
are the related attributes for table $d\in D_{cand}$. 
The second step translates the original OM query to SQL queries over the data tables.

In particular, for each candidate data table $d$,  
the translation process works in the following rules.

\begin{itemize}
\item The SELECT clause always get the distinct $d_{id}$ and/or
  record id $r_{id}$ (when it is needed);
\item The FROM clause has table $d \in D_{cand}$;
\item The WHERE clause contains all the function $f(\cdot)$ in the OM
  query;
\item When there is distinct constraint on entity or observation, the
  GROUP BY clause contains of all the attributes corresponding to the
  {\em key measurement types} of the entity or observation. we detail
  all process of getting the key measurement types of an observation
  later. 
\item The aggregation requirement in $f(\cdot)$ is translated to
  HAVING clause. 
\end{itemize}

Based on these rules, an OM query can be translated to the following
SQL scripts.

@@

The above describes the process for the OM query. 
In real application, people ask queries with distinct
observation or entity constraints. Also, people ask queries about
different observatitons using context. 
The following elaborate the details in processing these two cases.

%In detail,  how to deal with the distinct constraints.
When a query with distinct constraints, e.g., find data sets with {\em distinct} observations or
entities. To deal with this case, 
we need to be able to figure out what distinguishes one observation
from another. Here, one observation type's key measurement types can
play this role well. 
When an observation type is denoted with {\em distinct yes} and some
of its measurement types are denoted with {\em key yes}, 
the values of the measurements on the key measurement types can
uniquely distinguish one observation from the other. 
To implement this, in SQL, we can perform ``group by'' operations on
the key measurement types of the observations.

%In detail, how to deal with context
When the given OM query has context, we need to consider two cases. 
In the first case the context is not {\em identifying} to its
observation types, i.e., the context value
does not affect the uniqueness of one observation, this can be
processed as the basic way. 
When the context is denoted with {\em identifying yes}, it means the
identity of one observation also depend on the contextual
observation. 
In this case, we need to figure out the all the key measurement types
of one observation, then the ``GROUP BY'' operation should be on all
these key measurement types. 
\from{HP}{Put an example here.}
In this case, there is no need to change the SQL scripts.
the only change are the relevant attribute set $d_{a_1} \cdots
d_{a_m}$, which corresponds to not only the measurement types that
directly related to the observation, but also the measurement types
that are in the identifying context chain of the observation.

\from{HP}{Put a specific routine to get key measurements of
  observation and/or entity.}

\begin{example}\label{eg:query_reqerite_q1}
{\bf TODO: refine this example} 
Take $Q_1$ ($Observation: \langle Species='Picea~rubens'\rangle$) as
example. 
The {\tt OMconcept} is ``observation'' and {\tt cond} is ``{\em species = `Picea rubens'}'';

From the annotation $MT$, we can find the measurement type
$m_3$ with the characteristic {\em species}. 
From the mapping $Map$, we further find that the attribute {\em spp} that the
measurement type $m_3$ is mapped to.

The second step is to find the data sets which really have this value. 
For this one, we find the data table, and do a selection on the table
content.

\vspace{0.1in}
{\tt 
\begin{tabular}{l} 
SELECT DISTINCT $d_{id}$\\
FROM $d$\\
WHERE $spp = 'piru'$
\end{tabular}
}
\vspace{0.1in}

\end{example}

\begin{example}
Take $Q_3$ ($Observation:$\\
$\langle Species='Picea~rubens'$$\wedge count(distinct~Species)\geq
5\rangle$) as an example.

\from{HP}{Refine\\
For the first step, we find the data tables that contain the needed concepts. 
In each table, we need to figure out the distinct observations. 
From $m_i$ $obstype_i$, all the other measurements that are of
$obstype_i$. 
Group by the key measurements of $obstype_i$ if distinct yes.
If this observation type does not have any context, then the key
measurements are all the measurements directly defined under this
observatioin type. Otherwise, i.e., this observation type has context,
then the key measurements are all the measurements in the context
chain. }

\vspace{0.1in}
{\tt 
\begin{tabular}{l} 
SELECT DISTINCT $d_{id}$\\
FROM $d$\\
WHERE $spp = 'piru'$\\
GROUP BY $d_{id}$, $spp$\\
HAVING count($d_{id}$,$spp$) $\geq 5$\\
\end{tabular}
}
\vspace{0.1in}

\end{example}

%Analysis of this method. 
{\bf Cost analysis:} 
From the above description, we can tell that the computation cost is to search the different data
tables. This way, for each candidate data table, we need to send an
SQL query to the server for evaluation purpose.
If the number of candidate data tables are small. This method should
work well. However, when we have a lot of candidate data tables, the
efficiency may be affected. 
In what follows, we introduce another strategy, which logically merge
the data content in different tables and perform queries ober the
materialized database.

@@

The previous section shows the method to rewrite a given query. 
However, as we analyzed, the computation cost may be very high when
there are many candidate data tables. 
In this section, we propose another strategy to answer the given OM
query by making use of the existing optimization strategies of
current DBMS.

%rough idea of this strategy
In this strategy, we do not directly work on the original data table
$D$. On the other hand, we {\em materialize} the original data-sets to
centralized concept tables with instances of entities, observations, measurements
and contexts. Such concept tables are called core OBOE tables, and denoted as
$OBOE.*$. 
Then, the query process works on the materialized databases (i.e., the
conceptual data tables).

%structure of this sub-section
In what follows, we first describe the structure of the core
conceptual tables and present our algorithm to materialize the data into such
concept instance tables in Section \ref{sec:materialilzedb}. Our
algorithm considers all the distinct and context constraints. 
Then, Section \ref{sec:query_materialilzedb} provides the detailed
process of performing queries over materialized database.

\subsubsection{Materialize data}\label{sec:materialilzedb}

%Structure of conceptual instances
In particular, we have the following concept instance tables. 
\begin{itemize}
%\item $EI$ = $\{(\underline{ei_{id}}, et_{id})\}$ for all the entity instances;
\item $OI$ = $\{(\underline{oi_{id}}, ot_{id}, ei_{id})\}$ keeps
  all the observation instances where $ot_{id}$ is used to link to the
  annotation information in $OT$; 
\item $MI$ = $\{(\underline{mi_{id}}, oi_{id}, mt_{id},
  mVal)\}$ for all the measurement instances where $mt_{id}$ is used to link to the
  annotation information in $MT$; 
\item $CI =\{(\underline{oi_{id},coi_{id},ct_{id}})\}$ for
  all the context instances where $ct_{id}$ is used to link to the
  annotation information in $CT$. 
\end{itemize}

\from{HP}{TODO: refine the writing in MaterializeDB and move them here.}

@@

Once the data are materialized using the annotation constraints, we
can perform the OM query over them. 
%In what follows, we describe how to use the materialized data to answer the query.

@@

\vspace{0.1in}
{\tt 
\begin{tabular}{l}
SELECT distinct $AD.d_{id}$\\
FROM $AD, MI~as~mi, MT~as~mt$\\
WHERE $mt.Cha='Species'$ \\
$\qquad\quad$ AND $mi.mVal = `Picea~rubens'$\\
$\qquad\quad$ AND $mi.mt_{id}=mt.mt_{id}$\\
$\qquad\quad$ AND $mt.A_{id}=AD.A_{id}$\\
GROUP BY $d_{id},oi_{id}$\\
HAVING COUNT(*)$>$5; 
}
\end{tabular}
}
\vspace{0.1in}
\end{example}
